---
title: "fpashr: Continuous Density Estimation"
author: "Raphaël Brena, Tom McGrory, Kate Whelan, Conor Foran"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fpashr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(fpashr)
```

# Introduction


The `fpashr` package provides tools for continuous density estimation from numeric data. Four functions are provided:

  1. `compute_fp()`: Computes the quantities needed to construct a **frequency polygon (FP)** from numeric data by connecting histogram bin midpoints to form a continuous density estimate
  
  2. `compute_ash()`: Computes an **average shifted histogram (ASH)** density estimate by averaging multiple shifted histograms to reduce bin-edge bias
  
  3. `plot()`: An S3 plotting method for visualizing density estimates produced by continuous histogram estimators, including both frequency polygons and average shifted histograms
  
  4. `validate_density()`: Validates a density estimate by using numerical integration to check that the total area under the curve is approximately equal to 1

Usually, these functions are applied sequentially. This vignette aims to demonstrate a typical workflow of the `fpashr` package.


## Installation

__fpashr__ will run in Windows, Mac OS X, or Linux. To install it you first need to install [R](https://cran.r-project.org/). Installing [RStudio](https://www.rstudio.com/) as desktop environment for using R is also recommended.


Once in R you can type at the R command prompt:

```{r, eval=FALSE}
install.packages('devtools')
devtools::install_github('RaphaelBRENA/fpashr')
```

to install the latest version of the package from the __fpashr__ [GitHub page](https://github.com/RaphaelBRENA/fpashr). 

In either case, if you then type:

```{r, eval=FALSE}
library(fpashr)
```

it will load in all the __fpashr__ functions.

You will need to load in a numeric dataset, for e.g., the `geyser` dataset from `MASS`, which contains eruption data for the Old Faithful geyser in Yellowstone National Park.

```{r}
library(MASS)
data(geyser)
```


## `compute_fp()`

The `compute_fp()` function computes the values needed to build a **frequency polygon (FP)** from numeric data. A frequency polygon is formed by joining the midpoints of histogram bins, giving a smooth, continuous summary of the data.

The main input, `x`, is a numeric vector. The number of bins is set using `bins`, and `density` controls whether the output contains counts or densities (the default).

The function does not produce a plot. Instead, it returns the quantities needed for plotting or checking the estimate later.

```{r}
fp <- compute_fp(geyser$waiting, bins = 10)

#Midpoints
fp$x

#Counts/Densities
fp$y

#Histogram bin width
fp$binwidth
```
The output is an object, with class `"frequency_polygon"` and `"hist_estimate"`, containing the bin midpoints (`x`), the corresponding values (`y`), and the bin width (`binwidth`).


## `compute_ash()`

The `compute_ash()` function computes a density estimate using an **Average Shifted Histogram (ASH)**. It averages multiple histograms shifted by fractions of the bin width to reduce the bias caused by bin edges.
The main input, `x`, is a numeric vector. `h` sets the bin width for the underlying histograms, and `m` sets the number of shifts (default is 3).

Like `compute_fp()`, this function does not produce a plot. It returns the quantities needed for plotting or validation.



```{r}
ash <- compute_ash(geyser$waiting, h = 9, m = 3)

#Midpoints
ash$x

#Density values
ash$y

#Parameters used
#Bin width (h)
ash$parameters$h
#Number of shifts (m)
ash$parameters$m

```
The output is an object, with class `"ash_density"` and `"hist_estimate"`, containing the bin midpoints (`x`), the corresponding density values (`y`), and the parameters used (`h` for bin width and `m` for number of shifts).


## `plot()`

The `plot()` method displays a histogram overlaid with the density estimate for an object of class `"hist_estimate"`. It works with both frequency polygon (`compute_fp()`) and average shifted histogram (`compute_ash()`) objects.

```{r}
#Frequency polygon
fp <- compute_fp(geyser$waiting, bins = 10)
plot(fp)
#Average shifted histogram
ash <- compute_ash(geyser$waiting, h = 9, m = 3)
plot(ash)

```

The function returns a `ggplot` object, so you can modify it further. For example, you could change the title:

```{r}
library(ggplot2)
ash <- compute_ash(geyser$waiting, h = 9, m = 3)
p <- plot(ash, fill = "magenta", line = "purple")  +
ggtitle("ASH of Old Faithful Geyser Waiting Times")
p
```


## `validate_density()`

The `validate_density()` function checks that a density estimate integrates to approximately 1.0, as required for a valid probability density function. It uses numerical integration (the trapezoidal rule) to calculate the area under the curve.

The main input, `x`, is an object of class `"frequency_polygon"` or `"ash_density"` produced by `compute_fp()` or `compute_ash()`. The optional tolerance parameter sets the acceptable deviation from 1.0 (default 0.01).


```{r}
# Validate ASH density estimate
ash_result <- compute_ash(geyser$waiting, h = 9, m = 3)
validation_ash <- validate_density(ash_result)
print(validation_ash)

# Validate frequency polygon density estimate
fp_result <- compute_fp(geyser$waiting, bins = 15, density = TRUE)
validation_fp <- validate_density(fp_result)
print(validation_fp)

# Using a custom tolerance
validate_density(ash_result, tolerance = 0.0001)
```


The output is an object of class `"density_validation"` containing the total area under the density curve (`area`), the absolute deviation from 1.0 (`deviation`), the tolerance used (`tolerance`), and a logical flag indicating whether the density is valid (`valid`). If the density is not valid, printing the object will also display a warning message.
