---
title: "fpashr: Continuous Density Estimation"
author: "Raphaël Brena, Tom McGrory, Kate Whelan, Conor Foran"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fpashr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#library(fpashr)
devtools::load_all()
library(MASS)
library(ggplot2)
```

# Introduction
The `fpashr` package provides tools for continuous density estimation from numeric data. It includes:

- **Frequency Polygons (FP)** – connect histogram bin midpoints to create a continuous density curve.  
- **Average Shifted Histograms (ASH)** – reduce bin-edge bias by averaging multiple shifted histograms.  
- **Density Validation** – check that the estimated densities integrate to approximately 1 using `validate_density()`.

This vignette demonstrates how to:

1. Compute FP and ASH density estimates  
2. Visualize them with `plot.hist_estimate()`  
3. Validate the densities numerically with `validate_density()`

# Load the Example Data


